\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-01-21 jasa/paper-jasa.tex
Section 1: nitpicks

The likelihood function is the central object that summarizes the information
from an experiment needed for inference of model parameters. The likelihood
function is key to many areas of science that report the results of classical
hypothesis tests or confidence intervals using the (generalized or profile)
likelihood ratio as a test statistic. With the advance of computing technology,
it has become increasingly common that a simulator (or generative model) is used
to describe complex processes that tie parameters $\theta$ of an underlying
theory and measurement apparatus to high-dimensional observations $\mathbf{x}$. However,
directly evaluating the likelihood function in these cases is often impossible
or is computationally impractical. In this likelihood-free setting, various
methods for statistical inference have been proposed, including most notably the
Approximate Bayesian Computation (ABC) class of algorithms aiming at sampling
from the posterior distribution of model parameters \glnote{Add references}.

The main result of this paper is to show that the likelihood ratio is invariant
under dimensionality reduction, under the assumption that the transformation is
monotonic with the likelihood ratio. As a direct consequence, we derive and propose an
alternative machine learning-based approach for likelihood-free inference that
can also be used in a classical (frequentist) setting where a prior over the
model parameters is not available. More specifically, we demonstrate why and how
discriminative classifiers can be used to construct equivalent likelihood ratio
tests when only a generative model for the data is available for training and
calibration.

@@

As a concrete example, let us consider searches for new particles at the Large
Hadron Collider (LHC). The simulator that is sampling from $p(\mathbf{x}|\theta)$ is
based on quantum field theory, a detailed simulation of the particle detector,
and data processing algorithms that transform raw sensor data into the feature
vector $x$~\citep{Sjostrand:2006za,Agostinelli:2002hh}. The ATLAS and CMS
experiments have published  hundreds papers where the final result was
formulated as a hypothesis test or confidence interval using a generalized
likelihood ratio test~\citep{Cowan:2010js}, including most notably the discovery
of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa} and subsequent
measurement of its properties. The bulk of the likelihood ratio tests at the LHC
are based on the distribution of a single event-level feature that discriminates
between a hypothesized process of interest (labeled \textit{signal}) and various
other processes (labeled \textit{background}). Typically, pseudo-data from the
simulator are used to approximate the density at various parameter points, and
an interpolation algorithm is used to approximate the parametrized
model~\citep{Cranmer:2012sba}. In particular, to improve the statistical power
of these tests, hundreds of these searches have already been using supervised
learning to train discriminative classifiers that take advantage of a high
dimensional feature vector $\mathbf{x}$. \glnote{We should then elaborate on how the
proposed method could help these searches.}

@@

The rest of this paper is organized as follows. In Section...

>>>>

The likelihood function is the central object that summarizes the information
from an experiment needed for inference of model parameters. The likelihood
function is key to many areas of science that report the results of classical
hypothesis tests or confidence intervals using the (generalized or profile)
likelihood ratio as a test statistic. At the same time, with the advance of
computing technology, it has become increasingly common that a simulator (or
generative model) is used to describe complex processes that tie parameters
$\theta$ of an underlying theory and measurement apparatus to high-dimensional
observations $\mathbf{x}$. However, directly evaluating the likelihood function
in these cases is often impossible or is computationally impractical. In this
likelihood-free setting, various methods for statistical inference have been
proposed, including most notably the Approximate Bayesian Computation (ABC)
class of algorithms aiming at sampling from the posterior distribution of model
parameters \glnote{Add references}.

The main result of this paper is to show that the likelihood ratio is invariant
under dimensionality reduction, under the assumption that the corresponding
transformation is monotonic with the likelihood ratio. As a direct consequence,
we derive and propose an alternative machine learning-based approach for
likelihood-free inference that can also be used in a classical (frequentist)
setting where a prior over the model parameters is not available. More
specifically, we demonstrate why and how discriminative classifiers can be used
to construct equivalent likelihood ratio tests when only a generative model for
the data is available for training and calibration.

@@

As a concrete example, let us consider searches for new particles at the Large
Hadron Collider (LHC). The simulator that is sampling from $p(\mathbf{x}|\theta)$ is
based on quantum field theory, a detailed simulation of the particle detector,
and data processing algorithms that transform raw sensor data into the feature
vector $\mathbf{x}$~\citep{Sjostrand:2006za,Agostinelli:2002hh}. The ATLAS and CMS
experiments have published  hundreds papers where the final result was
formulated as a hypothesis test or confidence interval using a generalized
likelihood ratio test~\citep{Cowan:2010js}, including most notably the discovery
of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa} and subsequent
measurement of its properties. The bulk of the likelihood ratio tests at the LHC
are based on the distribution of a single event-level feature that discriminates
between a hypothesized process of interest (labeled \textit{signal}) and various
other processes (labeled \textit{background}). Typically, pseudo-data from the
simulator are used to approximate the density at various parameter points, and
an interpolation algorithm is used to approximate the parametrized
model~\citep{Cranmer:2012sba}. In particular, to improve the statistical power
of these tests, hundreds of these searches have already been using supervised
learning to train discriminative classifiers that take advantage of a high
dimensional feature vector $\mathbf{x}$. \glnote{We should then elaborate on how the
proposed method could help these searches.}

@@

The rest of this paper is organized as follows. In Section... \glnote{todo}
